---
title: "Homework 04: Data frames and data wrangling"
format: html
editor: visual
---

Please read the entire chapter on data transformation from\
[R for Data Science](https://r4ds.hadley.nz/data-transform) before starting this homework.

This homework relies on the `nycflights13` package, which contains several data frames, including:

-   `airlines`
-   `airports`
-   `flights`
-   `planes`
-   `weather`

Loading `nycflights13` puts these data frames on the search path.

------------------------------------------------------------------------

## Setup

### Load packages (do this every time)

```{r}
library(tidyverse)
library(nycflights13)
```

Note - if you don't have these installed you will need to first install.packages('nycflights13')

------------------------------------------------------------------------

# Question 1: Filtering

Make a plot of **air time vs distance** (air time on the y-axis, distance on the x-axis) for all flights that meet the following criteria:

-   originate from LaGuardia airport (`"LGA"`)
-   departed on the 16th of the month
-   have a flight distance of less than 2000 miles

## Your code

```{r}
# Filter flights according to the criteria above,
# then make the required plot.

flights_LGA <- flights |> 
  filter(distance<2000)|>
  filter(day == 16)|>
  filter(origin == "LGA")
  
ggplot(
  data = flights_LGA,
  mapping = aes(x = distance, y = air_time)
) +
  geom_point(na.rm = TRUE)
```

## Brief written response

In 2–4 sentences, describe what you observe in the plot. (For example: Is the relationship roughly linear? Are there any clear outliers?)

**Your answer here:** In the plot, we see a linear trend and no clear outliers. As the distance increases, we can observe that the air time also increases. There are some flights that have increased airtime for the same distance, which can indicate that the flight was delayed or that bad weather prevented the flight from landing on time.

------------------------------------------------------------------------

# Question 2: Dealing with NAs

Make a data frame of all rows of `flights` that have values for *both* `arr_time` and `dep_time` (i.e., neither value is `NA`).

## Your code

```{r}
# Create a new data frame that removes rows where arr_time or dep_time is NA.
flights_complete <- flights |>
  filter(!is.na(arr_time) & !is.na(dep_time))

```

## Filtering NAs (conceptual)

`ggplot()` will automatically remove `NA` values from a plot, but it emits a warning message about it. You *could* silence warnings using chunk options, but instead:

### Brief written response

Explain (in words) how you could prevent those `NA` values from appearing in the plot in the first place.

**Your answer here:** We can filter out the NA values by !is.na code that exclude all NA values and cleans up the row. Also, we can use the na.rm = (True) code to exclude in calculations.

------------------------------------------------------------------------

# Question 3: Adding columns

Create a data frame of **average flight speeds**, based on `air_time` and `distance`.

Then make either:

-   a histogram, **or**
-   a density plot

If you like, you may break the data out (e.g., by airline or another variable) in a way that you think makes sense.

## Your code

```{r}
# Create a new column for average speed.
# (Hint: think carefully about units — air_time is in minutes.)
# Then make a histogram or density plot.
average_speed <- flights |>
   filter(!is.na(air_time) & !is.na(distance)) |>
  mutate(avg_speed = distance / air_time * 60)
ggplot(average_speed, aes(x = avg_speed)) +
  geom_density(fill = "red", alpha = 0.5)
```

## Brief written response

Describe the main features of your speed distribution.

**Your answer here:** In the density plot, we can see that the avg. speed is clustering between 400-500 mph. This means the flights had an average speed of 400-500 mph.

------------------------------------------------------------------------

# Rendering and submission (GitHub)

Canvas contains the course GitHub link and instructions for **forking** the course repository.

In this file, your job is to:

1.  Complete the homework by adding code + written responses above.
2.  Render this Quarto file to **HTML or PDF**.
3.  Move the rendered file(s) to the /doc folder (note that, if you render to html you will also have to move your hmk_04_data_frames_files folder)
4.  Add/commit both:
    -   this edited `.qmd` file
    -   the rendered output file(s) (HTML or PDF)
5.  Push your changes to your fork on GitHub.
6.  Open a Pull Request back to the course repository.

## What to submit on Canvas

On Canvas, submit a link to your GitHub fork repository (not files).
